`SCPCL000824` - Using `SingleR` for classification of tumor cells

### 2024-07-01 analyses/cell-type-ewings/cell_type_annotations/SCPCS000492/SCPCL000824_singler-classification.Rmd
move heatmaps

@@

results_dir <- file.path(module_base, "results", "annotate_tumor_cells_output", "SCPCS000492")
geneset_scores_file <- file.path(results_dir, "SCPCL000824_gene-set-scores.tsv")

# singler results 
full_singler_results <- file.path(module_base, "scratch", "SingleR", "SCPCL000824_singler-results.rds")
singler_annotations_file <- file.path(annotations_dir, "SCPCL000824_singler-classifications.tsv")
```

@@

# read in annotations
annotations_df <- readr::read_tsv(annotations_file)
singler_annotations_df <- readr::read_tsv(singler_annotations_file)

@@

```{r}
# generate classification df
classification_df <- sce |> 
  scuttle::makePerCellDF(use.dimred = "UMAP") |> 
  # replace UMAP.1 with UMAP1
  dplyr::rename_with(
    \(x) stringr::str_replace(x, "^UMAP\\.", "UMAP")
  ) |>
  # get rid of excess columns
  dplyr::select(barcodes, UMAP1, UMAP2, singler_celltype_annotation) |> 
  dplyr::left_join(annotations_df, by = c("barcodes" = "cell_barcode")) |> 
  dplyr::left_join(singler_annotations_df, by = "barcodes") |> 
  dplyr::left_join(geneset_scores_df, by = "barcodes") |>
  # get the top cell types for each classification method for plotting later 
  dplyr::mutate(
    across(
      ends_with("_classification"),
      \(x) {
        x |>
          forcats::fct_lump_n(7, other_level = "All remaining cell types", ties.method = "first") |>
          forcats::fct_infreq() |>
          forcats::fct_relevel("All remaining cell types", after = Inf)
      },
      .names = "{.col}_lumped"
    )
  )
```

## SingleR

```{r}
annotations_to_plot <- c(
  "blueprint_tumor_classification_lumped",
  "updated_singler_classification_lumped",
  "updated_singler_blueprint_classification_lumped"
)

annotations_to_plot |> 
  purrr::map(\(annotation_column){
    ggplot(
    classification_df,
    aes(x = UMAP1, y = UMAP2, color = .data[[annotation_column]])
  ) +
    # set points for all "other" points
    geom_point(
      data = dplyr::select(
        classification_df, -.data[[annotation_column]]
      ),
      color = "gray80",
      alpha = 0.5,
      size = 0.5
    ) +
    # set points for desired cell type
    geom_point(size = 0.5, alpha = 0.5) +
    facet_wrap(
      vars(.data[[annotation_column]]),
      ncol = 3
    ) +
    scale_color_brewer(palette = "Dark2") +
    # remove axis numbers and background grid
    scale_x_continuous(labels = NULL, breaks = NULL) +
    scale_y_continuous(labels = NULL, breaks = NULL) +
    guides(
      color = guide_legend(
        title = "Cell types",
        # more visible points in legend
        override.aes = list(
          alpha = 1,
          size = 1.5
        )
      )
    )
  })
```

Compare tumor vs. other in all classification methods vs. updated

```{r}
plot_jaccard <- function(classification_df,
                         singler_results_column,
                         methods_to_compare
                         ){

  # create jaccard matrices for SingleR and CellAssign compared to aneuploid/diploid 
  jaccard_matrices <- methods_to_compare |>
    purrr::map(\(name) {
      make_jaccard_matrix(
        classification_df,
        singler_results_column,
        name
      )
    })

  # Set heatmap padding option
  heatmap_padding <- 0.2
  ComplexHeatmap::ht_opt(TITLE_PADDING = grid::unit(heatmap_padding, "in"))
  # list of heatmaps looking at SingleR/ CellAssign vs tumor/normal 
  heatmap <- jaccard_matrices |>
    purrr::imap(
      \(celltype_mat, celltype_method) {
        ComplexHeatmap::Heatmap(
          t(celltype_mat), # transpose because matrix rows are in common & we want a vertical arrangement
          col = circlize::colorRamp2(c(0, 1), colors = c("white", "darkslateblue")),
          border = TRUE,
          ## Row parameters
          cluster_rows = TRUE,
          row_title = celltype_method,
          row_title_gp = grid::gpar(fontsize = 12),
          row_title_side = "left",
          row_names_side = "left",
          row_dend_side = "right",
          row_names_gp = grid::gpar(fontsize = 10),
          ## Column parameters
          cluster_columns = TRUE,
          column_title = singler_results_column,
          column_title_gp = grid::gpar(fontsize = 12),
          column_names_side = "bottom",
          column_names_gp = grid::gpar(fontsize = 10),
          column_names_rot = 90,
          ## Legend parameters
          heatmap_legend_param = list(
            title = "Jaccard index",
            direction = "vertical",
            legend_width = unit(1.5, "in")
          ),
          show_heatmap_legend = celltype_method == "UCell",
        )
      }) |>
    # concatenate vertically into HeatmapList object
    purrr::reduce(ComplexHeatmap::`%v%`) |>
    ComplexHeatmap::draw(
      heatmap_legend_side = "right",
      # add a margin to the heatmap so labels don't get cut off
      padding = unit(c(2, 20, 2, 2), "mm")
    )

@@

@@

Look at consensus across all three references. 
For normal cells, identify some top markers and see if they are expressed in that group of cells compared to all other cells.

>>>>

This notebook looks at using `SingleR` to annotate tumor cells in `SCPCL000824`. 
We used three different references with `SingleR`:

1. Both the full `BlueprintEncodeData` reference from `celldex` and `SCPCL000822` where cells are either annotated as tumor or normal. 
Ambiguous cells are removed from `SCPCL000822`. 
2. `SCPCL000822` where all tumor cells are labeled as tumor and all other cells keep the original annotation obtained from running `SingleR` as part of `scpca-nf`.  
3. `SCPCL000822` with tumor cells labeled as in #2 combined with the full `BlueprintEncodeData` reference.

@@

# gene set scores to use for validation 
results_dir <- file.path(module_base, "results", "annotate_tumor_cells_output", "SCPCS000492")
geneset_scores_file <- file.path(results_dir, "SCPCL000824_gene-set-scores.tsv")

# singler results 
full_singler_results <- file.path(module_base, "scratch", "SingleR", "SCPCL000824_singler-results.rds")
singler_annotations_file <- file.path(annotations_dir, "SCPCL000824_singler-classifications.tsv.gz")
```

@@

# read in annotations
annotations_df <- readr::read_tsv(annotations_file)
singler_annotations_df <- readr::read_tsv(singler_annotations_file) |> unique()

@@

## Define functions

```{r}
# function to plot jaccard matrices to compare one set of annotations to other methods 
# `annotation_column` will be the columns and will be compared to each method listed in `methods_to_compare` 
plot_jaccard <- function(classification_df,
                         annotation_column, # single column of annotations to compare to all methods
                         methods_to_compare, # list of methods to compare to annotations
                         column_title, # title for columns
                         legend_match # what legend to keep, should be a name in `methods_to_compare`
){

  # create jaccard matrices 
  jaccard_matrices <- methods_to_compare |>
    purrr::map(\(name) {
      make_jaccard_matrix(
        classification_df,
        annotation_column,
        name
      )
    })

  heatmap <- make_heatmap_list(jaccard_matrices, column_title, legend_match)

@@

```{r}
# generate classification df to use for plots 
classification_df <- sce |> 
  scuttle::makePerCellDF(use.dimred = "UMAP") |> 
  # replace UMAP.1 with UMAP1
  dplyr::rename_with(
    \(x) stringr::str_replace(x, "^UMAP\\.", "UMAP")
  ) |>
  # get rid of excess columns
  dplyr::select(barcodes, UMAP1, UMAP2, singler_celltype_annotation) |> 
  # join with previous annotations, singler results, and gene set scores 
  dplyr::left_join(annotations_df, by = c("barcodes" = "cell_barcode")) |> 
  dplyr::left_join(singler_annotations_df, by = "barcodes") |> 
  dplyr::left_join(geneset_scores_df, by = "barcodes") |>
  # get the top cell types for each classification method for plotting later 
  dplyr::mutate(
    across(
      ends_with("_classification"),
      \(x) {
        x |>
          forcats::fct_lump_n(7, other_level = "All remaining cell types", ties.method = "first") |>
          forcats::fct_infreq() |>
          forcats::fct_relevel("All remaining cell types", after = Inf)
      },
      .names = "{.col}_lumped"
    )
  )
```

## SingleR results

### UMAPs of cell assignments

The first thing we will do is look at the annotated cell types on a UMAP. 
The top 7 cell types are shown for each reference used and all other cells are lumped together.

```{r}
# list of columns with singler results for UMAPs
singler_annotation_columns <- c(
  "blueprint_tumor_classification_lumped",
  "updated_singler_classification_lumped",
  "updated_singler_blueprint_classification_lumped"
)

singler_annotation_columns |> 
  purrr::map(\(annotation_column){
    ggplot(
    classification_df,
    aes(x = UMAP1, y = UMAP2, color = .data[[annotation_column]])
  ) +
    # set points for all "other" points
    geom_point(
      data = dplyr::select(
        classification_df, -.data[[annotation_column]]
      ),
      color = "gray80",
      alpha = 0.5,
      size = 0.5
    ) +
    # set points for desired cell type
    geom_point(size = 0.5, alpha = 0.5) +
    facet_wrap(
      vars(.data[[annotation_column]]),
      ncol = 3
    ) +
    scale_color_brewer(palette = "Dark2") +
    # remove axis numbers and background grid
    scale_x_continuous(labels = NULL, breaks = NULL) +
    scale_y_continuous(labels = NULL, breaks = NULL) +
    guides(
      color = guide_legend(
        title = "Cell types",
        # more visible points in legend
        override.aes = list(
          alpha = 1,
          size = 1.5
        )
      )
    )
  })
```

It appears that most cells in that bottom group of cells are being called tumor cells regardless of the reference that is used. 
The major differences here are in how the normal cells are labeled. 
For right now, let's focus on just identifying the tumor cells and see how similar each of the references are to each other before we compare to the other methods that we have used.

### Compare tumor cell annotations between references used

heatmap_list <- methods |>
  purrr::imap(\(method, name){

    plot_jaccard(classification_df,
                 annotation_column = method, # single column of annotations to compare to all methods
                 methods_to_compare = methods, # list of methods to compare to annotations
                 column_title = name,
                 legend_match = "Original" # what legend to keep, should be a name in `methods_to_compare`
    )

  }) 
heatmap_list
```

Upset plot looking at overlapping tumor cells called by all three

```{r}
tumor_cells <- list(
  "Blueprint" = classification_df$barcodes[classification_df$blueprint_tumor_classification == "Tumor"],
  "Original" = classification_df$barcodes[classification_df$updated_singler_classification == "Tumor"],
  "Original and Blueprint" = classification_df$barcodes[classification_df$updated_singler_blueprint_classification == "Tumor"]
)

UpSetR::upset(UpSetR::fromList(tumor_cells), order.by = "freq") |> print()
```

```{r}
classification_df <- classification_df |> 
  dplyr::mutate(singler_consensus_call = dplyr::case_when(
    blueprint_tumor_classification == "Tumor" & updated_singler_classification == "Tumor" & updated_singler_blueprint_classification == "Tumor" ~ "Tumor",
    blueprint_tumor_classification == "Tumor" | updated_singler_classification == "Tumor" | updated_singler_blueprint_classification == "Tumor" ~ "Ambiguous",
    .default = "Normal"
  ))

ggplot(classification_df, aes(x = UMAP1, y = UMAP2, color = singler_consensus_call)) +
  geom_point(alpha = 0.5, size = 0.5)
```

Compare normal calls to original SingleR normal calls 
```{r}
original_celltypes <- classification_df$singler_celltype_annotation |> unique()
normal_celltypes <- classification_df |> 
  dplyr::filter(updated_singler_blueprint_classification_lumped != "Tumor") |> 
  dplyr::pull(updated_singler_blueprint_classification_lumped) |> 
  unique()

shared_celltypes <- intersect(original_celltypes, normal_celltypes)

caret_df <- classification_df |> 
  dplyr::filter(singler_celltype_annotation %in% shared_celltypes |
                updated_singler_blueprint_classification %in% shared_celltypes)

caret::confusionMatrix(
    table(
      caret_df$singler_celltype_annotation, 
      caret_df$updated_singler_blueprint_classification)
  )s
```

### Compare tumor cell annotations with `SingleR` to marker gene based methods

@@

Look at consensus across all three references. 
For normal cells, identify some top markers and see if they are expressed in that group of cells compared to all other cells. 
confusion matrix across all shared normal cell types between OG singleR and new singleR
